---
title: 'R: The basics'
author: "Maggie Zhang - University of Illinois Urbana Champaign"
output:
  html_document:
    df_print: paged
---
# Introduction
This is a brief introduction to R for programming. 

The first step to use R is to set the working directory. Your working directory is the folder on your computer
in which you are currently working. When you ask R to open a file, it will look in the working directory for this file, and when you tell R to save a data file or figure, it will save it in the working directory.

```{r}
# setwd("/R Review") # set a working directory
```

I will recommend using the Project option when opening RStudio. You can either create a new project or connect a project with a pre-existing folder. Then you will be working in the project under the specified folder. 

R can do many statistical and data analyses. They are organized in so-called packages or libraries.
With the standard installation, most common packages are installed.

```{r}
#install.packages(dplyr)
library(dplyr)
# help(library = "dplyr)
```

A very useful function provided by dplyr package is the %>% pipe function. It can pass the value on the LHS to the function in the RHS 

```{r}
print("apple")
"apple" %>% print()
```


# Section I: Programming Basics
## 1.1. I/O
How to import data from text, csv files into R?
```{r}
#data = read.csv(file="https://drmaggiezhang.com/files/inclass_data.csv",stringsAsFactors=F)
#write.csv(data,file="new_data.csv")
```
If encountering new file types, Google it. The logic is the same.

### 1.1.0 RData 

Unique to R: save & load .RData files. 

1. using the Environment in RStudio 
2. using code


```{r}
# save(data,x,file="test.RData") # we can save multiple objects in one .RData file 
# load("test.RData")
# rm(x) # remove 
```


## 1.2. Basic syntax
Do some basic calculations. 
```{r}
1+2
1-2
1*2
1/2
sqrt(2) # squared root of 2
2^2 # 2 squared
```

We can define a variable and assign a value to it.

```{r}
height = 180 # instead of using the equation symbol, it is equivalent to use the arrow symbol.
weight = 50
print (height*weight)
```

or alternatively, we assign the result to a new variable

```{r}
nv = height*weight
nv # instead of print(nv)
```

R has some generic functions that we can use them directly. For example, calculating the sum and mean of a series of numbers.

```{r}
sum(c(1,2,3))
mean(c(1,2,3))
```

If you don't know how to use a function, you can use the help() function and check the examples. For example
```{r}
help(sum) # or equivalently
?sum
```

## 1.3. Data types
There are 3 general modes of data in R: string/characters, numbers, and logical.

```{r}
string = "I am xxx"
string

# use class or typeof to determine the type of a variable
class(string)
typeof(string)

number = 5
number

```
Another important data type is the logical type. There are two predefined variables:TRUE and FALSE
```{r}
logic = TRUE
logic
missing = NA
missing
```

## 1.4. Data structures
To learn any programming languages, a shortcut is to learn the core elements. Fortunately, most programming languages share very similar core elements. They are data structure, conditional expressions, loops, and functions. 

Four types of data structure are commonly used in R: vector, matrix, data frame, and list. The first is vector. Vector is a list of values. These values could be numeric, logic, or string.
```{r}
# using c (c is a function) to generate a new vector
v = c(1,7,9,10) 
v
```
The data type of vector values must be the same. If numbers,characters,and missing value are assigned together, what will happen? 
```{r}
v2 = c(1,7,NA,"hi")
v2
# to access an element in the vector
v[-2]
# vector elements can have names
names(v)
# assign names
names(v) = c("First","Second","Third","Fourth")
names(v)
```
We can apply mathematical calculations directly to vector. We can multiply a vector by a number, calculate the sqrt of each elements in the vector.

```{r}
v*2
sqrt(v)
```
The second structure is matrix. Matrix is a collection of vectors, organized in rows and columns. Values in a matrix should be in the same data type. Each column or row is a vector. We use the matrix function to generate a matrix with 5 cols and 5 rows. Values are integers from 1 to 25.

```{r}
m = matrix(1:25,5,5) 
```
Similar to vector, we could do simple maths.
```{r}
m+2
m+m
m%*%m # matrix nultiplication
```
We can combine two matrices by cols or rows.
```{r}
cbind(m,m) 
rbind(m,m)
# access a matrix element
m[2,3] = "hi"
```
A data frame is a matrix, while columns could be of all different data types. 

```{r}
v1=c(1,7,6,3,2)
v2=c(T,F,T,F,T)
v3=c("Y","N","Y","N","Y")
d = data.frame(v1,v2,v3,stringsAsFactors=F) # create use data.frame
```
A list is a collection of all data structure.
```{r}
l = list(v,m,d)
names(l)
l[[3]]
```

## 1.5. Programming tools
I will introduce 3 programming tools: if...else expression (which is one of the conditional expressions), for and while loop expressions, and user defined functions.

Task: input a number, if the value is larger than or equal to 100, print "High Value", if the value is less than 100 and larger than or equal to 50, print "Middle Value", if less than 50, print "Low Value".

```{r}
x = 95 
if (x>=100){
  print("High Value")
} else if (x>=50&x<100){
  print("Middle Value")
} else {
  print("Low Value")
}
```
Task: calculate the number of characters in our names. We use nchar() for each name.That means we will use the nchar() function many times.

```{r}
names=c("Hai","Jonanthan","Zhenzhen","Qinjie")
for (name in names) {
  N = nchar(name)
  print (paste(name,":",N,"characters"))
}
```
or alternatively, loop over the index of the vector:
```{r}
for (i in 1:4) {
  N = nchar(names[i])
  print (paste(names[i],":",N,"characters"))
}
```
or use while instead of loop

```{r}
i=1
while (i<=4){
  N = nchar(names[i])
  print (paste(names[i],":",N,"characters"))
  i=i+1 #counter
}
```
You can write your own functions and use them in the same way as the generic functions. For example, you can write a function to calculate the area of a square. The formula is area = width*height.

```{r}
area = function(height=5,width=20){
  area = height*width
  return (area)
}
area(11,10)
```
