---
title: "Lab 0 solutions"
author: "Maggie Zhang"
date: "2023-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Vector manipulation

For this section, you will be working on the following vector 

```{r}
L = c(3, 1, 1.3, 2, '2', c(2,8), TRUE, FALSE, FALSE, TRUE, 2, 3, 3, 9,2, "banana", "tree", 'bee', 'this is a long string')
```

### Q1. 
Write the code to show the data type of the 6th element in the vector L (Note: R uses the 1-based index).

```{r}
typeof(L[6])
```

Notice how R forced every element to be character type. 

### Q2. 
What is the data type of the 4th (2) and 5th element ('2') in the vector L respectively? 

```{r}
typeof(L[4])
```

```{r}
typeof(L[5])
```


### Q3. 
Find the total number of elements in the vector L. 

```{r}
length(L)
```

Notice how R will flatten everything. The 6th item, originally was a vector, but when put into the vector L as an element, it is forced to be flat and the 2 elements in it were released out. 

### Q4. 
Find the length (number of characters) of the last element in the vector L. 

```{r}
nchar(L[length(L)])
```
In R, there is no direct reverse indexing like the L[-1] in Python. But we can access the last element throught L[length(L)]. 

```{r}
L[-1]
```
Also notice the meaning of L[-1] in R: it will remove (-) the first element in the vector and return the rest. 


### Q5. 
Use vector slicing to access the 4 boolean values in the vector L. 

```{r}
L[c(8:11)]
```


### Q6. 
Print the vector in reverse order without changing it. 

```{r}
rev(L)
```


### Q7. 
Write code to calculate the sum of the first three numbers in the vector L. 

```{r}
sum(as.numeric(L[c(1:3)]))
```
First select the first three numbers by L[c(1:3)]. Then turn them into numbers by the as.numeric() function. Do the sum(). 

### Q8. 
Write code to count the frequency of 3 in the vector L. 

```{r}
table(L)
```
table() function in R can get the frequency table of items in the vector. 

### Q9. 
Add the following 2 elements to the vector L: “illini”, “spring”. 

```{r}
c(L,c("illini","spring"))
```


### Q10. 
Change the last element in the vector L to “this is a short string”.

```{r}
L[length(L)] = "this is a short string"
```


## Section 2: Data Frame 

In the following session, we will deal with a data frame. You can access the CSV file in the following link: https://drmaggiezhang.com/files/ElonMuskTweets.csv

This file contains all tweets posted by Elon Musk in the past 2 years from Jan/01/2021 to Jan/01/2023. I did the data collection using Twitter API. By the end of 1st module, you will also be able to do it as well. 

### Q11. 
Read in the CSV file into a data frame format. How many rows and columns of this data frame?

```{r}
df = read.csv("https://drmaggiezhang.com/files/ElonMuskTweets.csv",stringsAsFactors = F)
```

The stringAsFactors argument is to make sure when reading csv files, the strings won't be treated as a factor. 
 
### Q12. 
Find the most liked tweet with the maximum like_count. What is the id of that tweet?

```{r}
df[df$like_count == max(df$like_count),]
```


### Q13. 
What is the average retweet number of Elon Musk’s tweets?

```{r}
mean(df$retweet_count)
```


### Q14. 
What is the total length of tweet text?

```{r}
sum(nchar(df$text))
```


### Q15. 
Write a function that takes in 3 parameters as input: retweet_count, reply_count, and like_count. Calculate their average as the output. Name the function as CalculateEngagement 

```{r}
CalculateEngagement = function(retweet_count, reply_count, like_count){
  avg = (retweet_count + reply_count + like_count)/3
  return(avg)
}
```


### Q16. 
Using the CalculateEngagement functions you just wrote, create a new column in the data frame called engagement_index.

```{r}
df$engagement_index = CalculateEngagement(df$retweet_count, df$reply_count, df$like_count)
```


### Q17. 
Write another function that takes in 2 parameters: retweet_count, and reply_count. Calculate the reply-to-retweet ratio as the output (reply_count/retweet_count). Name the function as CalculateRatio.

```{r}
CalculateRatio = function(retweet_count, reply_count){
  ratio = reply_count/retweet_count
  return(ratio)
}
```


### Q18. 
Similarly, using the CalculateRatio functions you just wrote, create a new column in the data frame called ratio. Note: there are some tweets whose retweet_count is 0. Therefore you need to write a conditional statement. If the retweet_count is 0, set the ratio to 0. 

```{r}
ratio = c()
for (i in c(1:nrow(df))) {
  if (df$retweet_count[i] != 0){
    r = CalculateRatio(df$retweet_count[i], df$reply_count[i])
  }else{
    r = 0
  }
  ratio = c(ratio,r)
}
```

```{r}
df$ratio = ratio
```



### Q19. 
Get the full vector of all hashtags that appeared in Elon Musk’s tweets. (Note: you might need to combine the for loop and conditional statement, as well as the vector combination. The logic goes as follows: create an empty vector (named all_hashtags) to store all hashtags. For each item in the ElonMuskTweets[‘hashtags’], if it is empty, it means there is a vector of hashtags in that tweet. Append this vector to the all_hashtags vector you created. Use pandas.isna() to see if an element is empty). 



```{r}
all_hashtags = c()

for (hashtag in df$hashtags) {
  if (nchar(hashtag) != 0){
    all_hashtags = c(all_hashtags,hashtag)
  }
}
```

```{r}
all_hashtags
```


### Q20. 
Save the current data frame (with 2 additional columns) as a new CSV file. Upload it to Canvas. 

```{r}
write.csv(df,"results.csv")
```



